Map{R}educe in {MPI} for Large-scale Graph Algorithms

### 2010-11-19 paper/graph.tex
git-svn-id: svn+ssh://software-srn.sandia.gov/svn/mapreduce/trunk@672 3fc257bb-8550-0410-81b3-a5c103da83f7

We begin with a MapReduce procedure for creating large, sparse,
randomized graphs, since they are the input for the algorithms
discussed below.  R-MAT matrices ~\cite{RMAT} are recursively
generated graphs with power-law degree distributions.  They are
commonly used to represent web and social networks.  The user
specifices six parameters that define the graph: the number of
vertices $N$ and edges $M$, and four parameters $a$, $b$, $c$, $d$
that sum to 1.0 and are discussed below.  The algorithm in
Figure~\ref{fig:rmat} uses a single MapReduce object to generate $M$
unique non-zero entries in a sparse $N \times N$ matrix $G$, where
each entry $G_{ij}$ represents an edge between graph vertices
$(V_i,V_j)$, with $V_i$ an integer from $1$ to $N$.

@@

Input: \\
\> Pre-aggregated MapReduce object $A$ = $|V| \times |V|$ matrix:  Key = $i$, Value = $[j, A_{ij}]$ \\
\> Pre-aggregated MapReduce object $MT$ = indices of all-zero rows of $A$:  Key = $i$, Value = NULL \\
\> Pre-aggregated MapReduce object $x$ = initial PageRank vector:  Key = $i$, Value = $1/|V|$ \\
\> MapReduce object $y$ = vector, initially empty \\
\> User-defined parameter $\alpha$ \\
\\
while $({residual} > {tolerance})$ \\
\> 1 Compute adjustment for random jumps and zero-outdegree vertices \\
\> \> Add $x$ to $MT$ \\
$*$\> \> Convert $MT$:  Key is vertex $i$ \\
\> \> Reduce $MT$:   \\
\> \> \> Input:  Key = $i$, MultiValue = $(x_i, NULL)$ if $i \in {MT}$; else MultiValue = $x_i$ \\
\> \> \> If (${nvalues} = 2$) $c_{dot} = c_{dot} + x_i$; else $c_{dot} = c_{dot} + \alpha$ \\
\> \> \> Output: Key = $i$, Value = NULL if ${nvalues} = 2$ \\
\> \> $c_{dot}$ = MPI\_Allreduce($c_{dot}$, MPI\_SUM) / $|V|$ \\
\> 2 Compute $y = A^T x$  \\
\> \> Copy $x$ to $y$ \\
\> \> Add $A$ to $y$ \\
$*$\> \> Convert $y$:  Key is row index of $A$ \\
\> \> Reduce $y$: \\
\> \> \>  Input: Key = $i$, Multivalue = $y_i, [j, A_{ij}]$ for nonzeros $A_{ij}$ in $A$ \\
\> \> \>  Output: Key = $j$, Value = $A_{ij} y_i$ \\
\> \> Collate $y$:  Key is column index of $A$ \\
\> \> Reduce $y$: \\
\> \> \> Input: Key = $j$, Multivalue = $A_{ij} y_i$ for nonzeroes $A_{ij}$ in column $j$ of $A$ \\
\> \> \> Output: Key = $j$, Value = $\sum_i A_{ij} y_i$ \\
\> 3 Add $c_{dot}$ to $y$; compute $y_{max}$ = max element in $y$ \\
\> \> Map $y$:  \\
\> \> \> Input: Key = $i$, Value = $y_i$ \\
\> \> \> $y_i = y_i + c_{dot}$; $y_{max} = {\rm max}(y_i,y_{max})$ \\
\> \> \> Output:  Key = $i$, Value = $y_i$ \\
\> \> $y_{max}$ = MPI\_Allreduce($y_{max}$, MPI\_MAX) \\
\> 4 Scale $y$ by $y_{max}$ \\
\> \> Map $y$: \\
\> \> \> Input:  Key = $i$, Value = $y_i$ \\
\> \> \> Output:  Key = $i$, Value = $y_i / y_{max}$ \\
\> 5 Compute residual \\
\> \> Add $y$ to $x$ \\
$*$\> \> Convert $x$:  Key is vertex $i$ \\
\> \> Reduce $x$:  \\
\> \> \> Input: Key = $i$, Multivalue = $x_i, y_i$ \\
\> \> \> $residual = {\rm max}(|x_i - y_i|,residual)$ \\
\> \> $residual$ = MPI\_Allreduce($residual$, MPI\_MAX) \\
\> 6 Copy $y$ to $x$ \\
\\
Output: \\
\> MapReduce object $x$ = PageRank vector: Key = $i$, Value = $x_i$ = numeric rank

@@

Step 1 of a PageRank interation is, effectively, a dot-product of $MT$ and
$x$ to compute an adjustment $c_{dot}$ to the PageRank vector that
represents random jumps and the uniform probability of going from a 
leaf page to any
other page.  $MT$ is unchanged by the operation.  
In Step 2, the
matrix-vector product $A^T x$ is computed and the result stored in
$y$.  This is the most expensive part of the computation, requiring
two passes over the nonzero structure of $A$.  In the first pass, a
{\it convert()} gathers all local row entries $A_{ij}$ with their
associated $x_i$ entry, and a {\it reduce()} computes $A_{ij} x_i$.
In the second pass, a {\it collate()} gathers, for each $j$, all
contributions to the column sum $\sum_i A_{ij} x_i$, which is computed
by a second {\it reduce()}.  Steps 3 and 4 adjust and scale the
product vector; the work is proportional to $|V|$.  Step 5 computes
the residual as an infinity norm to determine whether the PageRank
vector has converged; work is again proportional to $|V|$.  Note that
$y$ and $x$ are both keyed by vertex at this point, so they can be
joined via an {\it add()} and {\it convert()} without need for the
communication required by a full {\it collate()}.  In Step 6, the
PageRank vector is overwritten prior to the next iteration.
Throughout the algorithm, we exploit the ability to call
MPI\_Allreduce to compute global dot products and norms.

@@

A connected component of a graph is a set of vertices where all pairs
of vertices in the set are connected by a path of edges.  A sparse
graph may contain many such components.  Figure~\ref{fig:cc} outlines
a MapReduce algorithm that labels each vertex in a graph with a
component ID.  All vertices in the same component are labelled with
the same ID, which is the ID of some vertex in the component.  We
assume an input graph $E$ of undirected edges $(V_i,V_j)$.  This
exposition also follows the connected-component algorithm presented
in~\cite{Cohen09}, with the addition of logic that load-balances data
across processors, which is important when one or a few giant
components exist in the graph.

@@

Input: \\
\> MapReduce object $E$ = edges of graph: Key = $(V_i,V_j)$, Value = NULL \\
\> MapReduce object $V$ = zone assignment of each vertex: Key = $V_i$, Value = $Z_i$ \\
\> MapReduce object $W$ = workspace
\\
Iterate: \\
\> 1 Map $W$ using $E$ as input: Convert edges to vertices \\
    \> \> Input: Key = $E_{ij} = (V_i, V_j)$, Value = NULL \\
    \> \> Output: Key = $V_i$, Value = $E_{ij}$ \\
    \> \> Output: Key = $V_j$, Value = $E_{ij}$ \\
\> 1 Add $V$ to $W$ \\
\> 1 Collate $W$: Vertex as key \\
\> 1 Reduce $W$: Emit edges of each vertex with zone of vertex \\
    \> \> Input: Key = $V_i$, MultiValue = $E E E E ... Z$ \\
    \> \> for each $E$ in MultiValue: \\
    \> \> \> Output: Key = $E_{ij}$, Value = $Z_i$ \\
\> 2 Collate $W$: Edge as key \\
\> 2 Reduce $W$: Emit zone re-assignments \\
       \> \> Input: Key = $E_{ij}$, MultiValue = $Z_i Z_j$ \\
       \> \> $Z_{winner}$ = min($Z_i$,$Z_j$); $Z_{loser}$ = max($Z_i$,$Z_j$) \\
       \> \> if $Z_i$ and $Z_j$ are different: \\
       \> \> \> Output: Key = $Z_{loser}$, Value = $Z_{winner}$ \\
\> 2 Exit: if $W$ is empty (no output by Reduce 2) \\
\> 3 Map $V$: Invert vertex/zone pairs \\
    \> \> Input: Key = $V_i$, Value = $Z_i$ \\
    \> \> if $Z_i$ is not partitioned: \\
    \> \> \> Output: Key = $Z_i$, Value = $V_i$ \\
    \> \> else: \\
    \> \> \> Output: Key = $Z_i^+$, Value = $V_i$ for a random processor \\
\> 4 Map $V$ using $W$ as input: Add zone reassignments in $W$ to $V$ \\
    \> \> if $Z_i$ is not partitioned: \\
    \> \> \> Output: Key = $Z_i$, Value = $Z_{winner}$ \\
    \> \> else: \\
    \> \> \> Output: Key = $Z_i^+$, Value = $Z_{winner}$ for every processor \\
    \> \> \> Output: Key = $Z_i$, Value = $Z_{winner}$ \\
\> 4 Collate $V$: Zone ID as key \\
\> 4 Reduce $V$: Emit new zone assignment of each vertex \\
       \> \> Input: Key = $Z_i$ or $Z_i^+$, MultiValue = $V V V V ... Z Z Z ...$ \\
       \> \> $Z_{new}$ = min($Z_i$ or $Z_i^+$,Z,Z,Z,...) \\
       \> \> partition $Z_{new}$ if number of $V >$ threshhold \\
       \> \> for each $V$ in MultiValue: \\
       \> \> \> Output: Key = $V_i$, Value = $Z_{new}$ \\
\\
Output: \\
\> MapReduce object $V$ = zone assignment of each vertex: Key = $V_i$, Value = $Z_i$ \\

@@

Note that if a graph has only a few components, the fourth {\it
collate()}, which keys on the zone ID, may generate a few very large
key/multivalue (KMV) pairs.  For example, if the entire graph is fully
connected, in the last iteration, a single KMV pair assigned to one
processor will result, containing all vertices in the graph.  This
imbalance in memory and computational work can lead to poor parallel
performance of the overall algorithm.  To counter this effect, various
operations in stages 3 and 4 include extra logic.  The idea is to
partition zones whose vertex count exceeds a user-defined threshhold
into $P$ sub-zones, where $P$ is the number of processors.  The 64-bit
integer that stores the zone ID also stores a bit flag indicating the
zone has been partitioned and a set of bits that encode the processor
ID.

@@

With this additional partitioning logic, the fourth {\it collate()}
(which keys on zone IDs, some of which now include processor bits)
collects only $1/P$ of the vertices in partitioned zones onto each
processor.  Yet the multivalue on each processor still contains all
the zone-reassignments relevant to the unpartitioned zone.  Thus, the
fourth {\it reduce()} can change the zone ID (if necessary) in a
consistent manner across all $P$ multivalues that contain the zone's
vertices.  When the fourth {\it reduce()} emits new zone assignments
for each vertex, the zone retains its partitioned status; the
partition bit is also explicitly set if the vertex count exceeds the
threshhold for the first time.

Note that this logic does not guarantee that the partition bits of the
zone IDs for all the vertices in a single zone are set consistently on
a given iteration.  For example, an unpartitioned zone with a small ID
may consume a partitioned zone.  The vertices from the partitioned
zone retain their partitioned status, but the original vertices in the
small zone may not set the partition bit of their zone IDs.  On
subsequent iterations, the fourth {\it map()} emits $P+1$ copies of
new zone reassignments for both partitioned and unpartioned zone IDs,
to ensure all vertices in the zone will know the correct reassignment
information.

@@

Input: \\
\> MapReduce object $V$ = maximal independent set vertices, initially empty \\
\> MapReduce object $E$ = edges of graph: Key = $(V_i,V_j)$ with $V_i < V_j$, Value = NULL \\
Map $E$: Assign random values consistently to each vertex $V_i$ and $V_j$ in $E$ \\
Clone $E$: Convert KV pairs in $E$ directly to KMV pairs with multivalue = NULL \\
\\
while $N_{edges}$ in $E > 0$: \\
\> 1 Reduce $E$: Determine WINNER/LOSER vertex of each edge \\
       \> \> Input: Key = $E_{ij}$, Multivalue = NULL or LOSER \\
       \> \> if multivalue = LOSER, emit nothing \\
       \> \> $V_w$ = WINNER vertex, $V_l$ = LOSER vertex \\
       \> \> Output: Key = $V_w$, Value = ($V_l$, WINNER) \\
       \> \> Output: Key = $V_l$, Value = ($V_w$, LOSER) \\
\> 1 Collate $E$: Vertex as key \\
\> 2 Reduce $E$: Find WINNER vertices \\
       \> \> Input: Key = $V_i$, MultiValue = $V V V V ...$ \\
       \> \> if all $V$ are flagged with WINNER, $V_i$ is a WINNER \\
       \> \> \> for each $V_j$ in Multivalue: \\
       \> \> \> \> Output: Key = $V_j$, Value = ($V_i$, WINNER) \\
       \> \> else: \\
       \> \> \> for each $V_j$ in Multivalue: \\
       \> \> \> \> Output: Key = $V_j$, Value = $V_i$ \\
\> 2 Collate $E$: Vertex as key \\
\> 3 Reduce $E$: Find LOSER vertices \\
       \> \> Input: Key = $V_i$, MultiValue = $V V V V ...$ \\
       \> \> if any $V$ is flagged with WINNER, $V_i$ is a LOSER \\
       \> \> \> for each $V_j$ in Multivalue: \\
       \> \> \> \> Output: Key = $V_j$, Value = ($V_i$, LOSER) \\
       \> \> else: \\
       \> \> \> for each $V_j$ in Multivalue: \\
       \> \> \> \> Output: Key = $V_j$, Value = $V_i$ \\
\> 3 Collate $E$: Vertex as key \\
\> 4 Reduce $E$: Emit WINNER vertices and LOSER edges \\
       \> \> Input: Key = $V_i$, MultiValue = $V V V V ...$ \\
       \> \> if all $V$ are flagged with LOSER, $V_i$ is a WINNER \\
       \> \> \> Output: to $V$: Key = $V_i$, Value = NULL \\
       \> \> for each $V_j$ in Multivalue: \\
       \> \> \> if $V_j$ is flagged with LOSER: \\
       \> \> \> \> Output: to $E$: Key = $E_{ij}$ with $V_i < V_j$, Value = LOSER \\
       \> \> \> else: \\
       \> \> \> \> Output: to $E$: Key = $E_{ij}$ with $V_i < V_j$, Value = NULL \\
\> 4 Collate $E$: Edge as key \\
\\
Output: \\
\> MapReduce object $V$ = maximal independent set vertices: Key = $V_i$, Value = NULL \\

@@

In the third {\it reduce()}, if any neighbor of $V_i$ was a winner in
the previoue {\it reduce()} (indicated by a winner flag), $V_i$
becomes a ``loser'' vertex and emits all its neighbor vertices with
$V_i$ tagged with a loser flag.  Otherwise it emits all its neighbor
vertices without the loser flag on $V_i$.

@@

For a distributed graph, global information about edge adjacencies and
labeled vertices is not available.  Instead, following the
Bellman-Ford-style algorithm of~\cite{SSSPMapReduce, Bellman58,
Ford62}, every edge of the graph is visited in each iteration and
vertex distancs are updated.  A MapReduce formulation of this
algorithm, described in Figure~\ref{fig:sssp}, begins with a graph $E$
of directed edges $(V_i,V_j)$.  As generated by the R-MAT procedure
disussed earlier, this may include self-edges $(V_i,V_i)$ and loop
edges (both $(V_i,V_j)$ and $(V_j,V_i)$).  These are handled as
special cases within the algorithm to avoid spurious contributions to
the shortest path distances.  MapReduce object $V$ stores the distance
of each vertex $V_i$ from the source vertex $V_s$.

@@

Input: \\
\> MapReduce object $E$ = graph edges with weights: Key = ($V_i, V_j$), Value = $w_{ij}$ \\
\> MapReduce object $V$ = vertices with distances from source vertex $V_s$: \\
\> \> Key = $V_i$, Value = $\infty$, except Value for $V_s$ = 0 \\
\\
while not $done$: \\
\> $done$ = 1 \\
\> Map $V$ using $E$ as input: Add edge weights in $E$ to $V$ \\
\> \> Input: Key = $E_{ij}$, Value = $w_{ij}$ \\
\> \> Output: Key = $V_i$, Value = $(V_j, w_{ij})$ \\
\> Collate $V$: Vertex is key \\
\> Reduce $V$:  Loop over edges of vertex to update distance to $V_s$ \\
\> \> Input: Key = $V_i$, Multivalue = $d_i$, edges $(V_j, w_{ij})$, candidate distances $d_j$ \\
\> \> $d_{min} = {\rm min}(d_i,d_j,d_j,\ldots)$ \\
\> \> If $d_{min} \ne d_i$: \\
\> \> \>  $done$ = 0 \\
\> \> \>  Output: Key = $V_j$, Value = $d_{min} + w_{ij}$ for each edge vertex $V_j$ \\
\> \>  Output:  Key = $V_i$, Value = $d_{min}$ \\
\> $done$ = MPI\_Allreduce($done$,MPI\_MIN) \\
\\
Output: \\
\> MapReduce object $V$ = vertices with distances from source vertex $V_s$: \\
\> \> Key = $V_i$, Value = $d_i$ \\

>>>>

We begin with a MapReduce procedure for creating large, sparse,
randomized graphs, since they are the input for the algorithms
discussed below.  R-MAT matrices ~\cite{RMAT} are recursively
generated graphs with power-law degree distributions.  They are
commonly used to represent web and social networks.  The user
specifies six parameters that define the graph: the number of
vertices $N$ and edges $M$, and four parameters $a$, $b$, $c$, $d$
that sum to 1.0 and are discussed below.  The algorithm in
Figure~\ref{fig:rmat} uses a single MapReduce object to generate $M$
unique non-zero entries in a sparse $N \times N$ matrix $G$, where
each entry $G_{ij}$ represents an edge between graph vertices
$(V_i,V_j)$, with $V_i$ an integer from $1$ to $N$.

@@

Input: \\
\> Pre-aggregated MapReduce object $A$ = $|V| \times |V|$ matrix:  Key = $i$, Value = $[j, A_{ij}]$ \\
\> Pre-aggregated MapReduce object $MT$ = indices of all-zero rows of $A$:  Key = $i$, Value = NULL \\
\> Pre-aggregated MapReduce object $x$ = initial PageRank vector:  Key = $i$, Value = $1/|V|$ \\
\> MapReduce object $y$ = vector, initially empty \\
\> User-defined parameter $\alpha$ \\
\\
while $({residual} > {tolerance})$ \\
\> 1 Compute adjustment for random jumps and zero-outdegree vertices \\
\> \> Add $x$ to $MT$ \\
$*$\> \> Convert $MT$:  Key is vertex $i$ \\
\> \> Reduce $MT$:   \\
\> \> \> Input:  Key = $i$, MultiValue = $(x_i, NULL)$ if $i \in {MT}$; else MultiValue = $x_i$ \\
\> \> \> If (${nvalues} = 2$) $c_{dot} = c_{dot} + x_i$; else $c_{dot} = c_{dot} + \alpha$ \\
\> \> \> Output: Key = $i$, Value = NULL if ${nvalues} = 2$ \\
\> \> $c_{dot}$ = MPI\_Allreduce($c_{dot}$, MPI\_SUM) / $|V|$ \\
\> 2 Compute $y = A^T x$  \\
\> \> Copy $x$ to $y$ \\
\> \> Add $A$ to $y$ \\
$*$\> \> Convert $y$:  Key is row index of $A$ \\
\> \> Reduce $y$: \\
\> \> \>  Input: Key = $i$, MultiValue = $y_i, [j, A_{ij}]$ for nonzeros $A_{ij}$ in $A$ \\
\> \> \>  Output: Key = $j$, Value = $A_{ij} y_i$ \\
\> \> Collate $y$:  Key is column index of $A$ \\
\> \> Reduce $y$: \\
\> \> \> Input: Key = $j$, MultiValue = $A_{ij} y_i$ for nonzeroes $A_{ij}$ in column $j$ of $A$ \\
\> \> \> Output: Key = $j$, Value = $\sum_i A_{ij} y_i$ \\
\> 3 Add $c_{dot}$ to $y$; compute $y_{max}$ = max element in $y$ \\
\> \> Map $y$:  \\
\> \> \> Input: Key = $i$, Value = $y_i$ \\
\> \> \> $y_i = y_i + c_{dot}$; $y_{max} = {\rm max}(y_i,y_{max})$ \\
\> \> \> Output:  Key = $i$, Value = $y_i$ \\
\> \> $y_{max}$ = MPI\_Allreduce($y_{max}$, MPI\_MAX) \\
\> 4 Scale $y$ by $y_{max}$ \\
\> \> Map $y$: \\
\> \> \> Input:  Key = $i$, Value = $y_i$ \\
\> \> \> Output:  Key = $i$, Value = $y_i / y_{max}$ \\
\> 5 Compute residual \\
\> \> Add $y$ to $x$ \\
$*$\> \> Convert $x$:  Key is vertex $i$ \\
\> \> Reduce $x$:  \\
\> \> \> Input: Key = $i$, MultiValue = $x_i, y_i$ \\
\> \> \> $residual = {\rm max}(|x_i - y_i|,residual)$ \\
\> \> $residual$ = MPI\_Allreduce($residual$, MPI\_MAX) \\
\> 6 Copy $y$ to $x$ \\
\\
Output: \\
\> MapReduce object $x$ = PageRank vector: Key = $i$, Value = $x_i$ = numeric rank

@@

Step 1 of a PageRank iteration is, effectively, a dot-product of $MT$ and
$x$ to compute an adjustment $c_{dot}$ to the PageRank vector that
represents random jumps and the uniform probability of going from a 
leaf page to any
other page.  $MT$ is unchanged by the operation.  
In Step 2, the
matrix-vector product $A^T x$ is computed and the result stored in
$y$.  This is the most expensive part of the computation, requiring
two passes over the nonzero structure of $A$.  In the first pass, a
{\it convert()} gathers all local row entries $A_{ij}$ with their
associated $x_i$ entry, and a {\it reduce()} computes $A_{ij} x_i$.
In the second pass, a {\it collate()} gathers, for each $j$, all
contributions to the column sum $\sum_i A_{ij} x_i$, which is computed
by a second {\it reduce()}.  Steps 3 and 4 adjust and scale the
product vector; the work is proportional to $|V|$.  Step 5 computes
the residual as an infinity norm to determine whether the PageRank
vector has converged; work is again proportional to $|V|$.  Note that
$y$ and $x$ are both keyed by vertex at this point, so they can be
joined via an {\it add()} and {\it convert()} without need for the
communication required by a full {\it collate()}.  In Step 6, the
PageRank vector is overwritten prior to the next iteration.
Throughout the algorithm, we exploit the ability to call
MPI\_Allreduce to compute global dot products and norms.

@@

A connected component of a graph is a set of vertices where all pairs
of vertices in the set are connected by a path of edges.  A sparse
graph may contain many such components.  Figure~\ref{fig:cc} outlines
a MapReduce algorithm that labels each vertex in a graph with a
component ID.  All vertices in the same component are labeled with
the same ID, which is the ID of some vertex in the component.  We
assume an input graph $E$ of undirected edges $(V_i,V_j)$.  This
exposition also follows the connected-component algorithm presented
in~\cite{Cohen09}, with the addition of logic that load-balances data
across processors, which is important when one or a few giant
components exist in the graph.

@@

Input: \\
\> MapReduce object $E$ = edges of graph: Key = $(V_i,V_j)$, Value = NULL \\
\> MapReduce object $V$ = zone assignment of each vertex: Key = $V_i$, Value = $Z_i$ \\
\> MapReduce object $W$ = workspace
\\
Iterate: \\
\> 1 Map $W$ using $E$ as input: Convert edges to vertices \\
    \> \> Input: Key = $E_{ij} = (V_i, V_j)$, Value = NULL \\
    \> \> Output: Key = $V_i$, Value = $E_{ij}$ \\
    \> \> Output: Key = $V_j$, Value = $E_{ij}$ \\
\> 1 Add $V$ to $W$ \\
\> 1 Collate $W$: Vertex as key \\
\> 1 Reduce $W$: Emit edges of each vertex with zone of vertex \\
    \> \> Input: Key = $V_i$, MultiValue = $E E E E ... Z$ \\
    \> \> for each $E$ in MultiValue: \\
    \> \> \> Output: Key = $E_{ij}$, Value = $Z_i$ \\
\> 2 Collate $W$: Edge as key \\
\> 2 Reduce $W$: Emit zone re-assignments \\
       \> \> Input: Key = $E_{ij}$, MultiValue = $Z_i Z_j$ \\
       \> \> $Z_{winner}$ = min($Z_i$,$Z_j$); $Z_{loser}$ = max($Z_i$,$Z_j$) \\
       \> \> if $Z_i$ and $Z_j$ are different: \\
       \> \> \> Output: Key = $Z_{loser}$, Value = $Z_{winner}$ \\
\> 2 Exit: if $W$ is empty (no output by Reduce 2) \\
\> 3 Map $V$: Invert vertex/zone pairs \\
    \> \> Input: Key = $V_i$, Value = $Z_i$ \\
    \> \> if $Z_i$ is not partitioned: \\
    \> \> \> Output: Key = $Z_i$, Value = $V_i$ \\
    \> \> else: \\
    \> \> \> Output: Key = $Z_i^+$, Value = $V_i$ for a random processor \\
\> 4 Map $V$ using $W$ as input: Add zone reassignments in $W$ to $V$ \\
    \> \> if $Z_i$ is not partitioned: \\
    \> \> \> Output: Key = $Z_i$, Value = $Z_{winner}$ \\
    \> \> else: \\
    \> \> \> Output: Key = $Z_i^+$, Value = $Z_{winner}$ for every processor \\
    \> \> \> Output: Key = $Z_i$, Value = $Z_{winner}$ \\
\> 4 Collate $V$: Zone ID as key \\
\> 4 Reduce $V$: Emit new zone assignment of each vertex \\
       \> \> Input: Key = $Z_i$ or $Z_i^+$, MultiValue = $V V V V ... Z Z Z ...$ \\
       \> \> $Z_{new}$ = min($Z_i$ or $Z_i^+$,Z,Z,Z,...) \\
       \> \> partition $Z_{new}$ if number of $V >$ threshold \\
       \> \> for each $V$ in MultiValue: \\
       \> \> \> Output: Key = $V_i$, Value = $Z_{new}$ \\
\\
Output: \\
\> MapReduce object $V$ = zone assignment of each vertex: Key = $V_i$, Value = $Z_i$ \\

@@

Note that if a graph has only a few components, the fourth {\it
collate()}, which keys on the zone ID, may generate a few very large
key/multivalue (KMV) pairs.  For example, if the entire graph is fully
connected, in the last iteration, a single KMV pair assigned to one
processor will result, containing all vertices in the graph.  This
imbalance in memory and computational work can lead to poor parallel
performance of the overall algorithm.  To counter this effect, various
operations in stages 3 and 4 include extra logic.  The idea is to
partition zones whose vertex count exceeds a user-defined threshold
into $P$ sub-zones, where $P$ is the number of processors.  The 64-bit
integer that stores the zone ID also stores a bit flag indicating the
zone has been partitioned and a set of bits that encode the processor
ID.

@@

With this additional partitioning logic, the fourth {\it collate()}
(which keys on zone IDs, some of which now include processor bits)
collects only $1/P$ of the vertices in partitioned zones onto each
processor.  Yet the multivalue on each processor still contains all
the zone-reassignments relevant to the unpartitioned zone.  Thus, the
fourth {\it reduce()} can change the zone ID (if necessary) in a
consistent manner across all $P$ multivalues that contain the zone's
vertices.  When the fourth {\it reduce()} emits new zone assignments
for each vertex, the zone retains its partitioned status; the
partition bit is also explicitly set if the vertex count exceeds the
threshold for the first time.

Note that this logic does not guarantee that the partition bits of the
zone IDs for all the vertices in a single zone are set consistently on
a given iteration.  For example, an unpartitioned zone with a small ID
may consume a partitioned zone.  The vertices from the partitioned
zone retain their partitioned status, but the original vertices in the
small zone may not set the partition bit of their zone IDs.  On
subsequent iterations, the fourth {\it map()} emits $P+1$ copies of
new zone reassignments for both partitioned and unpartitioned zone
IDs, to ensure all vertices in the zone will know the correct
reassignment information.

@@

Input: \\
\> MapReduce object $V$ = maximal independent set vertices, initially empty \\
\> MapReduce object $E$ = edges of graph: Key = $(V_i,V_j)$ with $V_i < V_j$, Value = NULL \\
Map $E$: Assign random values consistently to each vertex $V_i$ and $V_j$ in $E$ \\
Clone $E$: Convert KV pairs in $E$ directly to KMV pairs with multivalue = NULL \\
\\
while $N_{edges}$ in $E > 0$: \\
\> 1 Reduce $E$: Determine WINNER/LOSER vertex of each edge \\
       \> \> Input: Key = $E_{ij}$, MultiValue = NULL or LOSER \\
       \> \> if multivalue = LOSER, emit nothing \\
       \> \> $V_w$ = WINNER vertex, $V_l$ = LOSER vertex \\
       \> \> Output: Key = $V_w$, Value = ($V_l$, WINNER) \\
       \> \> Output: Key = $V_l$, Value = ($V_w$, LOSER) \\
\> 1 Collate $E$: Vertex as key \\
\> 2 Reduce $E$: Find WINNER vertices \\
       \> \> Input: Key = $V_i$, MultiValue = $V V V V ...$ \\
       \> \> if all $V$ are flagged with WINNER, $V_i$ is a WINNER \\
       \> \> \> for each $V_j$ in MultiValue: \\
       \> \> \> \> Output: Key = $V_j$, Value = ($V_i$, WINNER) \\
       \> \> else: \\
       \> \> \> for each $V_j$ in MultiValue: \\
       \> \> \> \> Output: Key = $V_j$, Value = $V_i$ \\
\> 2 Collate $E$: Vertex as key \\
\> 3 Reduce $E$: Find LOSER vertices \\
       \> \> Input: Key = $V_i$, MultiValue = $V V V V ...$ \\
       \> \> if any $V$ is flagged with WINNER, $V_i$ is a LOSER \\
       \> \> \> for each $V_j$ in MultiValue: \\
       \> \> \> \> Output: Key = $V_j$, Value = ($V_i$, LOSER) \\
       \> \> else: \\
       \> \> \> for each $V_j$ in MultiValue: \\
       \> \> \> \> Output: Key = $V_j$, Value = $V_i$ \\
\> 3 Collate $E$: Vertex as key \\
\> 4 Reduce $E$: Emit WINNER vertices and LOSER edges \\
       \> \> Input: Key = $V_i$, MultiValue = $V V V V ...$ \\
       \> \> if all $V$ are flagged with LOSER, $V_i$ is a WINNER \\
       \> \> \> Output: to $V$: Key = $V_i$, Value = NULL \\
       \> \> for each $V_j$ in MultiValue: \\
       \> \> \> if $V_j$ is flagged with LOSER: \\
       \> \> \> \> Output: to $E$: Key = $E_{ij}$ with $V_i < V_j$, Value = LOSER \\
       \> \> \> else: \\
       \> \> \> \> Output: to $E$: Key = $E_{ij}$ with $V_i < V_j$, Value = NULL \\
\> 4 Collate $E$: Edge as key \\
\\
Output: \\
\> MapReduce object $V$ = maximal independent set vertices: Key = $V_i$, Value = NULL \\

@@

In the third {\it reduce()}, if any neighbor of $V_i$ was a winner in
the previous {\it reduce()} (indicated by a winner flag), $V_i$
becomes a ``loser'' vertex and emits all its neighbor vertices with
$V_i$ tagged with a loser flag.  Otherwise it emits all its neighbor
vertices without the loser flag on $V_i$.

@@

For a distributed graph, global information about edge adjacencies and
labeled vertices is not available.  Instead, following the
Bellman-Ford-style algorithm of~\cite{SSSPMapReduce, Bellman58,
Ford62}, every edge of the graph is visited in each iteration and
vertex distance are updated.  A MapReduce formulation of this
algorithm, described in Figure~\ref{fig:sssp}, begins with a graph $E$
of directed edges $(V_i,V_j)$.  As generated by the R-MAT procedure
discussed earlier, this may include self-edges $(V_i,V_i)$ and loop
edges (both $(V_i,V_j)$ and $(V_j,V_i)$).  These are handled as
special cases within the algorithm to avoid spurious contributions to
the shortest path distances.  MapReduce object $V$ stores the distance
of each vertex $V_i$ from the source vertex $V_s$.

@@

Input: \\
\> MapReduce object $E$ = graph edges with weights: Key = ($V_i, V_j$), Value = $w_{ij}$ \\
\> MapReduce object $V$ = vertices with distances from source vertex $V_s$: \\
\> \> Key = $V_i$, Value = $\infty$, except Value for $V_s$ = 0 \\
\\
while not $done$: \\
\> $done$ = 1 \\
\> Map $V$ using $E$ as input: Add edge weights in $E$ to $V$ \\
\> \> Input: Key = $E_{ij}$, Value = $w_{ij}$ \\
\> \> Output: Key = $V_i$, Value = $(V_j, w_{ij})$ \\
\> Collate $V$: Vertex is key \\
\> Reduce $V$:  Loop over edges of vertex to update distance to $V_s$ \\
\> \> Input: Key = $V_i$, MultiValue = $d_i$, edges $(V_j, w_{ij})$, candidate distances $d_j$ \\
\> \> $d_{min} = {\rm min}(d_i,d_j,d_j,\ldots)$ \\
\> \> If $d_{min} \ne d_i$: \\
\> \> \>  $done$ = 0 \\
\> \> \>  Output: Key = $V_j$, Value = $d_{min} + w_{ij}$ for each edge vertex $V_j$ \\
\> \>  Output:  Key = $V_i$, Value = $d_{min}$ \\
\> $done$ = MPI\_Allreduce($done$,MPI\_MIN) \\
\\
Output: \\
\> MapReduce object $V$ = vertices with distances from source vertex $V_s$: \\
\> \> Key = $V_i$, Value = $d_i$ \\
